Evaluating infectious disease forecasts with allocation scoring rules

### 2023-12-19 alloscore_manuscript/alloscore-application-shorter.Rnw

\paragraph{Example 1} Suppose we have a forecast $F$ for need in two locations with $F_1 = \mathrm{Exp}(1 / \sigma_1)$ and $F_2 = \mathrm{Exp}(1 / \sigma_2)$, where $\sigma_1 = 1$ and $\sigma_2 = 4$. The means of these distributions are given by the scale parameters $\sigma_i$. When the marginal forecasts are exponential distributions, it can be shown that the optimal allocation divides the available resources among the locations proportionally to the scale parameters $\sigma_i$ (see section
\ref{sec:bayes-quantiles} of the supplemental materials). If $K = 5$ units of the resource are available, the optimal allocation according to $F$ would be 1 unit of resources in location 1 and 4 units of resources in location 2. If, on the other hand, $K = 10$ units are available, we will allocate 2 units of resources to location 1 and 8 units to location 2. Figure~\ref{fig:exp_alloc_example} illustrates the situation.

@@

left_join(ias, as) |>
  left_join(wis_vals, by="model") |>
  rename(Model = model,
         `IAS centered at 15k` = IAS_wt,
         `IAS uniform` = IAS) |>
  select(Model, AS, MWIS, `IAS centered at 15k`, `IAS uniform`) |>
  arrange(AS) |>
  xtable::xtable(caption = "Comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.",
                 digits = 0,
                 label = "tab:multi-k-scores") |>
  print(include.rownames = FALSE, )
@

@@

<<multi-loc-ranks, fig.width=8, fig.cap="Component-wise breakdowns of the allocation score (Panel A) and weighted interval score (Panel B), by location for forecasts of hospitalization admissions on January 3, 2022, for two selected models (JHUAPL-Bucky and CU-select). Panel A shows the observed resource need, in this case the observed number of hospitalizations, for each state, along with the hypothetical number of resources allocated to the given location based on the forecasts from each model. The number of available resources was fixed at 15,000 and forecasts from each model were used to determine an optimal allocation strategy before the resource need was known. For most locations the resource need exceeded the resources allocated, indicated by the `observed resource need' bar being larger than the `allocation' bar. Panel B shows the breakdown of the weighted interval score (WIS) into components of underprediction, overprediction and dispersion. Larger values of WIS indicate more error, and the full WIS score for each location can be decomposed into the three components shown here.">>=
## date points to target_end_date of 2022-01-10 to highlight
ref_date_for_plot <- "2021-12-20"
models_to_plot <- c("JHUAPL-Bucky", "CU-select")

@@

\subsubsection{Forecast scores showed differences over time}
<<grob-hosps-over-time>>=
dat_to_plot <- truth_data |>
  filter(location == "US",
         target_end_date %in% score_data_summaries$target_end_date)
p1 <- ggplot(dat_to_plot,
             aes(x=target_end_date, y=value)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=15000, linetype=2) +
  ylab("Hospital admissions") +
  xlab(NULL) +
  ylim(c(0, NA))
@

<<metrics-over-time, fig.height=8, fig.width=8, fig.cap="Hospital admissions and evaluation metrics over time. Panel A shows the number of hospital admissions in the US as a whole due to COVID-19 on a sequence of 15 Mondays from December 2021 through March 2022. These are the dates for which forecasts were made and evaluated. A horizontal dashed line at 15,000 shows the assumed resource constraint $K$. Panel B shows allocation scores (AS) for each model's 14 day-ahead forecast, across all US states. The x-axis corresponds to the date for which the prediction was made. AS typically are high when the observed value is near to the constraint, which occurs during the last Monday in December (on the way up) and the last Monday in January (on the way down). Panel C shows the MWIS metric across weeks, averaged across all states. MWIS values tend to scale with the observed and predicted values, and the peak MWIS values happen around and just after the peak of the Omicron wave.">>=
p2 <- ggplot(alloscore_data_summaries,
             aes(x=target_end_date, y=alloscore, color=model, shape=model)) +
  scale_x_date(limits=range(score_data_summaries$target_end_date)) +
  scale_shape_manual(values=c(15:18, 4, 6, 15:18, 4, 6, 15:18, 4)) +
  ylab("Allocation score") +
  geom_point() + geom_line()+
  theme(legend.pos="none") +
  xlab(NULL)

p4 <- ggplot(score_data_summaries,
             aes(x=target_end_date, y=mwis, color=model,shape=model)) +
  geom_point() + geom_line() +
  scale_shape_manual(values=c(15:18, 4, 6, 15:18, 4, 6, 15:18, 4)) +
  xlab("Forecast target date") +
  ylab("MWIS")+
  theme(legend.pos="bottom")+
  guides(colour = guide_legend(title=NULL, ncol = 4),
         shape = guide_legend(title=NULL, ncol = 4))

cowplot::plot_grid(p1, p2, p4,
                   nrow=3,
                   align="v",
                   rel_heights = c(1, 1, 1.85),
                   labels = "AUTO")
@

@@

<<metrics-correlation, fig.width=8, fig.height=9, fig.cap="Association of standardized ranks for MWIS and allocation score by model and week. Each facet of the plot corresponds to one model. Within each facet, each point corresponds to a week. The x- and y-values correspond to the MWIS standardized rank and the allocation score standardized rank for that week. Points corresponding to earlier dates have darker shading. The size of the point corresponds to the observed value on the date for which the prediction was made. Models show different degrees of association between the two metrics.">>=
all_scores |>
  filter(as.Date(reference_date.x) < as.Date("2022-02-01") ) |>
  dplyr::mutate(target_end_date = as.Date(target_end_date)) |>
  ggplot(aes(x=mwis_rank_std, y=allo_rank_std, color=target_end_date, size=yobs)) +
    geom_point(alpha=.7) +
    scale_x_continuous("MWIS standardized rank", limits=c(0,1)) +
    scale_y_continuous("Allocation score standardized rank", limits=c(0,1)) +
#    scale_color_gradient("Target date") +
    scale_color_date(
      name="Target date",
      breaks = seq.Date(from=as.Date("2021-12-13"), by = 14, length.out=5)) +
    scale_size_continuous("Observed\nhospitalizations") +
    facet_wrap(.~model, ncol=3) +
    geom_abline(slope=1, intercept=0, linetype=2, color="gray") #+
#    theme(legend.pos="bottom")

@@

p1 <- ggplot(ias_to_plot, aes(x=AS, y=IAS, color=IAS_type, shape=IAS_type)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE, linetype=2) +
  scale_color_manual(name = "Integrated allocation score weighting", values = c("#386cb0", "#fdc086")) +
  scale_shape_manual(name = "Integrated allocation score weighting", values = c(1, 2)) +
  geom_abline(slope=1, intercept=0, color="gray") +
  xlab("Allocation score") +
  ylab("Integrated allocation score") +
  theme(legend.position = c(0.05,.9), legend.justification = c(0,1),
        legend.background=element_rect(fill = alpha("white", 0.2)))

p2 <- slim_alloscore_2021.12.20 |>
  group_by(model, K) |>
  summarize(`Allocation score` = sum(components)) |>
  left_join(kprobs) |>
  ggplot(aes(x=K)) +
  geom_point(aes(y=weight), color = "#386cb0") +
  geom_point(y=1/length(unique(slim_alloscore_2021.12.20$K)), color="#fdc086", shape = 2) +
  geom_vline(xintercept = 15000, color="gray") +
  ylab("Weight")

>>>>

\paragraph{Example 1} Suppose we have a forecast $F$ for need in two locations with $F_1 = \mathrm{Exp}(1 / \sigma_1)$ and $F_2 = \mathrm{Exp}(1 / \sigma_2)$, where $\sigma_1 = 1$ and $\sigma_2 = 4$. When the marginal forecasts are exponential distributions, it can be shown that the optimal allocation divides the available resources among the locations proportionally to the scale parameters $\sigma_i$ (see section
\ref{sec:bayes-quantiles} of the supplemental materials). If $K = 5$ units of the resource are available, the optimal allocation according to $F$ would be 1 unit of resources in location 1 and 4 units of resources in location 2. If, on the other hand, $K = 10$ units are available, we will allocate 2 units of resources to location 1 and 8 units to location 2. Figure~\ref{fig:exp_alloc_example} illustrates the situation.

@@

left_join(ias, as) |>
  left_join(wis_vals, by="model") |>
  rename(`IAS centered at 15k` = IAS_wt,
         `IAS uniform` = IAS) |>
  select(model, AS, MWIS, `IAS centered at 15k`, `IAS uniform`) |>
  arrange(AS) |>
  xtable::xtable(caption = "Comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.",
                 digits = 0,
                 label = "tab:multi-k-scores") |>
  print(include.rownames = FALSE, )
@

@@

<<multi-loc-ranks, fig.cap="Component-wise breakdowns of the allocation score (Panel A) and weighted interval score (Panel B), by location for forecasts of hospitalization admissions on January 3, 2022, for two selected models (JHUAPL-Bucky and CU-select). Panel A shows the observed resource need, in this case the observed number of hospitalizations, for each state, along with the hypothetical number of resources allocated to the given location based on the forecasts from each model. The number of available resources was fixed at 15,000 and forecasts from each model were used to determine an optimal allocation strategy before the resource need was known. For most locations the resource need exceeded the resources allocated, indicated by the `observed resource need' bar being larger than the `allocation' bar. Panel B shows the breakdown of the weighted interval score (WIS) into components of underprediction, overprediction and dispersion. Larger values of WIS indicate more error, and the full WIS score for each location can be decomposed into the three components shown here.">>=
## date points to target_end_date of 2022-01-10 to highlight
ref_date_for_plot <- "2021-12-20"
models_to_plot <- c("JHUAPL-Bucky", "CU-select")

@@

\subsubsection{Forecast scores showed differences over time}
<<grob-hosps-over-time>>=
dat_to_plot <- truth_data |>
  filter(location == "US",
         target_end_date %in% score_data_summaries$target_end_date)
p1 <- ggplot(dat_to_plot,
             aes(x=target_end_date, y=value)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept=15000, linetype=2) +
  ylab("hospital admissions") +
  xlab(NULL) +
  ylim(c(0, NA))
@

<<metrics-over-time, fig.height=9, fig.width=7, fig.cap="Hospital admissions and evaluation metrics over time. Panel A shows the number of hospital admissions in the US as a whole due to COVID-19 on a sequence of 15 Mondays from December 2021 through March 2022. These are the dates for which forecasts were made and evaluated. A horizontal dashed line at 15,000 shows the assumed resource constraint $K$. Panel B shows allocation scores (AS) for each model's 14 day-ahead forecast, across all US states. The x-axis corresponds to the date for which the prediction was made. AS typically are high when the observed value is near to the constraint, which occurs during the last Monday in December (on the way up) and the last Monday in January (on the way down). Panel C shows the MWIS metric across weeks, averaged across all states. MWIS values tend to scale with the observed and predicted values, and the peak MWIS values happen around and just after the peak of the Omicron wave.">>=
p2 <- ggplot(alloscore_data_summaries,
             aes(x=target_end_date, y=alloscore, color=model, shape=model)) +
  scale_x_date(limits=range(score_data_summaries$target_end_date)) +
  scale_shape_manual(values=c(15:18, 4, 6, 15:18, 4, 6, 15:18, 4)) +
  ylab("Allocation score") +
  geom_point() + geom_line()+
  theme(legend.pos="none") +
  xlab(NULL)

p4 <- ggplot(score_data_summaries,
             aes(x=target_end_date, y=mwis, color=model,shape=model)) +
  geom_point() + geom_line() +
  scale_shape_manual(values=c(15:18, 4, 6, 15:18, 4, 6, 15:18, 4)) +
  ylab("MWIS")+
  theme(legend.pos="bottom")+
  xlab(NULL)

cowplot::plot_grid(p1, p2, p4,
                   nrow=3,
                   align="v",
                   rel_heights = c(1, 1, 1.5),
                   labels = "AUTO")
@

@@

<<metrics-correlation, fig.width=10, fig.height=7.5, fig.cap="Association of standardized ranks for MWIS and allocation score by model and week. Each facet of the plot corresponds to one model. Within each facet, each point corresponds to a week. The x- and y-values correspond to the MWIS standardized rank and the allocation score standardized rank for that week. Points corresponding to earlier dates have darker shading. The size of the point corresponds to the observed value on the date for which the prediction was made. Models show different degrees of association between the two metrics.">>=
all_scores |>
  filter(as.Date(reference_date.x) < as.Date("2022-02-01") ) |>
  ggplot(aes(x=mwis_rank_std, y=allo_rank_std, color=target_end_date, size=yobs)) +
  geom_point(alpha=.7) +
  scale_x_continuous("MWIS standardized rank", limits=c(0,1)) +
  scale_y_continuous("Allocation score standardized rank", limits=c(0,1)) +
  facet_wrap(.~model) +
  geom_abline(slope=1, intercept=0, linetype=2, color="gray")

@@

p1 <- ggplot(ias_to_plot, aes(x=AS, y=IAS, color=IAS_type, shape=IAS_type)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE, linetype=2) +
  scale_color_manual(name = "Integrated Allocation Score", values = c("#386cb0", "#fdc086")) +
  scale_shape_manual(name = "Integrated Allocation Score", values = c(1, 2)) +
  geom_abline(slope=1, intercept=0, color="gray") +
  theme(legend.position = c(0.05,.9), legend.justification = c(0,1),
        legend.background=element_rect(fill = alpha("white", 0.2)))

p2 <- slim_alloscore_2021.12.20 |>
  group_by(model, K) |>
  summarize(`Allocation score` = sum(components)) |>
  left_join(kprobs) |>
  ggplot(aes(x=K)) +
  geom_point(aes(y=weight), color = "#386cb0") +
  geom_point(y=1/length(unique(slim_alloscore_2021.12.20$K)), color="#fdc086", shape = 2) +
  geom_vline(xintercept = 15000, color="gray")
